Supplement to ``{\mytitle}''

### 2021-09-02 si/si.Rnw
code to make unformatted accepted version

<<debug,echo=F>>=
#run_level <- 1
#run_level <- "jasa-sept-2020"
run_level <- 2
abfNbhd_run_level <- run_level
lz_run_level <- run_level
bm_run_level <- run_level
mscale_run_level <- run_level
slice_run_level <- run_level

if(run_level=="jasa-sept-2020"){  
#### values used for jasa submission, sept 2020  
  abfNbhd_run_level <- 3
  lz_run_level <- 5
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
}

@@

\input{header-ms.tex}
\input{theorems.tex}

@@

\renewcommand{\contentsname}{Supplementary Content}
\renewcommand{\refname}{Supplementary References}
\renewcommand\thefigure{S-\arabic{figure}}
\renewcommand\thetable{S-\arabic{table}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\theprop{S\arabic{prop}}
\renewcommand\thelemma{S\arabic{lemma}}

@@

\begin{center}
\noindent\begin{tabular}{l}
\hline
{\bf Algorithm~G-{\UBF} (Generalized unadapted bagged filter).}\rule[-1.5mm]{0mm}{6mm}\\
\hline
{\bf input:}\rule[-1.5mm]{0mm}{6mm} \\
Simulator for $f_{{X}_{\time}|{X}_{\time-1}}({x}_{\time}|{x}_{\time-1})$\\
Evaluator for $f_{Y_{\unit,\time}|{X}_{\time}}(\data{y}_{\unit,\time}\given {x}_{\time})$\\
Number of bootstrap filters, $\Rep$\\
Neighborhood structure, $B_{\unit,\time}$, for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$\\
Data, $\data{y}_{\unit,\time}$ for $\unit\in 1\mycolon\Unit$ and $\time\in 1\mycolon\Time$ \\
{\bf output:}\rule[-1.5mm]{0mm}{5mm} \\
Log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit,\time}$  \rule[-2mm]{0mm}{3mm}
\\
\hline
For $\rep$ in $1\mycolon\Rep$ \rule[-1.5mm]{0mm}{6mm}
\\
\asp simulate ${X}_{\time,\rep}^{\simulation}$ from the dynamic model, for $\time\in 1{\mycolon}\Time$\\
End For\\
Prediction weights, 
$w^P_{\unit,\time,\rep}= f_{Y_{B_{\unit,\time}}|{X}_{1:\time}}(\data{y}_{B_{\unit,\time}}\given {X}^{\simulation}_{1:\time,\rep})$\\
Measurement weights,
$w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\time},Y_{B_{\unit,\time}}}(\data{y}_{\unit,\time}\given X^{\simulation}_{\time,\rep},\data{y}_{B_{\unit,\time}})$
\\
Conditional log likelihood estimate,
$\MC{\loglik}_{\unit,\time}= 
\log \left(
  \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}\,   w^P_{\unit,\time,\rep}
\right)
-
\log \left(
    \sum_{\tilde \rep=1}^\Rep w^P_{\unit,\time,\tilde \rep}
\right)
$
\\
\hline
\end{tabular}
\end{center}

@@

Call the adapted simulation process $\{A_n\}$, defined conditionally on $\{Y_n\}$. 
We see from the above calculation that 

@@

$\{A_n\}$ tracks $\{X_n\}$ when the process $\{A_n-X_n\}$ is stable.

@@

Together, the results in [\ref{thm1:loglik:linearization}], [\ref{var:bound:without:assumption}], [\ref{var:bound:with:assumption}] and [\ref{thm1:bias:bound}] confirm the assertions of the theorem.

@@

The mixing of the adapted process in Assumption~\ref{B:unconditional:mix} replaces the mixing of the unconditional process in  Assumption~\ref{A:unconditional:mix}.
Though mixing of the adapted process may be hard to check, one may suspect that the adapted process typically mixes more rapidly than the unconditional process.
%As for the TIF analysis, we note that requiring the inequality in Assumption~\ref{B:unconditional:mix} to hold over ${B^{+2c}_{\unit,\time}}$ implies that it holds over all subsets of  ${B^{+2c}_{\unit,\time}}$. 
Assumption~\ref{B:unconditional:mix} is needed only to ensure that the variance bound in Theorem~\ref{thm:abf} is essentially $O(\Unit \Time)$ rather than $O(\Unit^2 \Time^2)$. 
Either of these rates avoids the exponentially increasing variance characterizing the curse of dimensionality.
Lower variance than $O(\Unit \Time)$ cannot be anticipated for any sequential Monte Carlo method since the log likelihood estimate can be written as a sum of $\Unit\Time$ terms each of which involves its own sequential Monte Carlo calculation.
The following Proposition~\ref{prop:cov:mix} gives an implication of Assumption~\ref{B:unconditional:mix}

@@

bm_files_dir <- paste0("bm_",bm_run_level,"/")
if(!dir.exists(bm_files_dir)) dir.create(bm_files_dir)

@@

measles_model_dir <- "measlesModel/"
load(file=paste0(measles_model_dir,"measles_spatPomp.rda"))

@@

mscale_files_dir <- paste0("mscale_",mscale_run_level,"/")

@@

\begin{table}
\renewcommand{\arraystretch}{1.2}
  \centering
  \begin{tabular}{|crcl|}
    \hline
    parameter & value & unit & description  \\  \hline
    $\meanBeta$ & \Sexpr{myround(measles_params["R0"]*(measles_params["gamma"]+measles_params["mu"]),1)} & year$^{-1}$& mean contact rate\\
    $\mu_{\bullet D}^{-1}$ & \Sexpr{myround(1/measles_params["mu"],1)} &year& mean duration in the population \\
    $\mu_{EI}^{-1}$ & \Sexpr{myround(365/measles_params["sigma"],1)} &day& latent period  \\
    $\mu_{IR}^{-1}$ & \Sexpr{myround(365/measles_params["gamma"],1)} &day& infectious period \\
    $\sigma_{SE}$ & \Sexpr{myround(measles_params["sigmaSE"],3)} &year$^{1/2}$& process noise\\
    $\amplitude$ & \Sexpr{myround(measles_params["amplitude"],3)} &--- & amplitude of seasonality  \\
    $\alpha$ & \Sexpr{measles_params["alpha"]} &---& mixing exponent \\
    $\tau$ & 4 & year & delay from birth to entry into susceptibles\\    
%%    $\cohort$ & \Sexpr{measles_params["amplitude"]} &---& cohort fraction entering $S$ on September 6 \\
    $\rho$ & \Sexpr{measles_params["rho"]} &---& reporting probability \\
    $\psi$ & \Sexpr{measles_params["psi"]} &---& reporting overdispersion \\
    $G$ & \Sexpr{measles_params["g"]} &---& gravitation constant \\  \hline
  \end{tabular}
  \caption{\label{tab:measles_parameters} Table of parameters for the spatiotemporal measles transmission model}
\end{table}

@@

Table~\ref{tab:measles_parameters} gives the model parameter values and Table~\ref{tab:mscale} gives the algorithmic settings used for the filters.
The times in Table~\ref{tab:mscale} give the total time required by each algorithm to calculate all its results for Figure~{\MainFigureMeaslesScaling} in the main text, using \Sexpr{mscale_cores} cores.
The expected forecast function $\mu(x,s,t)$ needed for {\ABFIR} and GIRF was computed using a numerical solution to the deterministic skeleton of the stochastic model, i.e, a system of ODEs with derivative matching the infinitesimal mean function of the stochastic dynamic model.
In the specifications of $h_\unit\big(x)$, $\VtoTheta_\unit(V,x,\theta)$ and $\thetaToV_\unit(\theta,x)$, the latent process value $x$ contains a variable $C$ giving the cumulative removed infections in the current observation interval.

In Table~\ref{tab:mscale}, we see that the effort allocated to {\UBF}, {\ABF} and PF scales linearly with $\Unit$, since the number of bootstrap replications and particles is fixed in this experiment. 
GIRF computational effort scales fairly linearly in $\Unit$, since its effort is dominated by the guide simulations (which are linear in $\Unit$) rather than by the intermediate timestep calculations (which are quadratic in $\Unit$ since we carry out $\Unit$ intermediate calculations each of size linear in $\Unit$).
The effort allocated to {\ABFIR} scales with $\Unit^2$, since {\ABFIR} is more parsimonious with guide simulations (all particles in one bootstrap replication share the same guide simulations) and so the intermediate timestep calculations dominate the effort.
To obtain stable variance in the log likelihood estimate, the number of particles and bootstrap replications would have to grow with $\Unit$.
However, given a constraint on total computational resources, the number of particles and bootstrap replications would have to shrink as $\Unit$ increases.
The limit studied in this experiment is a balance between the two: the assumption is that one is prepared to invest a growing amount of computational effort as the data grow, but this should not grow too fast.
{\ABFIR} was permitted the greatest computational effort, but two considerations balance this.
\begin{enumerate}
\item Parallelization. {\UBF}, {\ABF} and {\ABFIR} are trivially parallelizable. 
The value of parallelization depends, among other things, on how many replications are being computed simultaneously and on how many cores are available.
Nevertheless, it is helpful that the core minute effort requirement for {\ABF} and {\ABFIR} can be divided by the number of available cores to give the computational time. 
Parallelizations of GIRF and PF can be constructed \citep{park20} but these involve non-trivial interaction between processors leading to additional algorithmic complexity and computational overhead. 
\item Memory. 
The intermediate timestep calculations in {\ABFIR} and GIRF do not add to the memory requirement, and the memory demands of {\UBF}, {\ABF} and {\ABFIR} are distributed across the parallel computations. 
A basic PF implementation for a large model can become constrained by its memory requirement (linear in the number of particles) before it can match the processor effort employed by the other algorithms.
\end{enumerate}

@@

  abfNbhd_model_dir <- "measlesModel/"
  # note: care is required if abfNbhd_model_dir is set to something other than the default measlesModel
  # however, it may be useful to do that while testing model variations.

@@

\section{\secTitleSpace A Lorenz-96 example}

@@

  set.seed(4512)
  lz_list <- foreach(u=lz_U) %do% {
    lz1 <- lorenz(
      U=u, N=lz_N,
      dt=lz_dt,dt_obs=lz_dt_obs,
      regular_params=lz_params)
    lz_ivps <- rnorm(u,mean=lz_ivp_mean,sd=lz_ivp_sd)
    names(lz_ivps) <- paste0("X",1:u,"_0")
    lz1_params <- coef(lz1)
    lz1_params[names(lz_ivps)] <- lz_ivps
    simulate(lz1,params=lz1_params)
  }
})

@@

<<lz_girf,cache=F,echo=F>>=
lz_girf <- stew(file=paste0(lz_files_dir,"lz_girf.rda"),seed=5981724,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      girf(lz_list[[job$U_id]], method='adams',
        Np=lz_girf_Np,
        Ninter = job$U,
        lookahead = lz_girf_lookahead,
        Nguide = lz_girf_nguide,
        tol = lz_tol) -> lz_girf_out 
    ) -> lz_girf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"],lz_girf_out)
    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"])
  } -> lz_girf_list
})
lz_jobs$girf_logLik <- vapply(lz_girf_list,function(x)x$logLik,numeric(1))
lz_jobs$girf_time <- vapply(lz_girf_list,function(x) x$time,numeric(1))
@

@@

<<lz_ubf2,cache=F,echo=F,eval=F>>=
lz_ubf2 <- stew(file=paste0(lz_files_dir,"lz_ubf2.rda"),seed=844424,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abf(lz_list[[job$U_id]],
        Nrep = lz_ubf_Nrep,
        Np=1,
        nbhd = lz_nbhd2, tol=lz_tol) -> lz_ubf2_out
    ) -> lz_ubf2_time
#    uncomment for debugging
#    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"],lz_ubf2_out)
    list(logLik=logLik(lz_ubf2_out),time=lz_ubf2_time["elapsed"])
  } -> lz_ubf2_list
})
lz_jobs$ubf2_logLik <- vapply(lz_ubf2_list,function(x)x$logLik,numeric(1))
lz_jobs$ubf2_time <- vapply(lz_ubf2_list,function(x) x$time,numeric(1))

<<lz_abfir,cache=F,echo=F,eval=T>>=
lz_abfir <- stew(file=paste0(lz_files_dir,"lz_abfir.rda"),seed=53398,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abfir(lz_list[[job$U_id]], method='adams',
        Nrep = as.integer(lz_abfir_Nrep),
        Np=lz_abfir_Np_per_replicate,
        Ninter = as.integer(job$U/2),
        nbhd = lz_nbhd, tol=lz_tol) -> lz_abfir_out 
    ) -> lz_abfir_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"],lz_abfir_out)
    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"])
  } -> lz_abfir_list
})
lz_jobs$abfir_logLik <- vapply(lz_abfir_list,function(x)x$logLik,numeric(1))
lz_jobs$abfir_time <- vapply(lz_abfir_list,function(x) x$time,numeric(1))
@

@@

lz_enkf <- stew(file=paste0(lz_files_dir,"lz_enkf.rda"),seed=837509,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      enkf(lz_list[[job$U_id]], 
        Np=lz_enkf_Np,
        h = lz_h,
        R = diag((coef(lz_list[[job$U_id]])["tau"])^2, nrow = job$U)  
      ) -> lz_enkf_out 
    ) -> lz_enkf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"],lz_enkf_out)
    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"])
  } -> lz_enkf_list
})
lz_jobs$enkf_logLik <- vapply(lz_enkf_list,function(x)x$logLik,numeric(1))
lz_jobs$enkf_time <- vapply(lz_enkf_list,function(x) x$time,numeric(1))

@@

<<lz_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('Lorenz \'96 simulation used in the main text')>>=
library(fields)
lz_obj <- lz_list[[which.max(lz_U)]]
par(mai=c(0.7,0.7,0.1,0.5))
image.plot(
  y=1:dim(states(lz_obj))[1],
  x=time(lz_obj),z=t(states(lz_obj)),
  ylab="unit",xlab="time",col=gray.colors(33)
)
@

@@

\bibliography{bib-iif}

>>>>

<<debug,echo=F>>=
#run_level <- 1
#run_level <- "jasa-sept-2020"
 run_level <- "manual"
#run_level <- 2

if(run_level=="manual"){
  abfNbhd_run_level <- 3
  lz_run_level <- 5
  ij_run_level <- 2
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
  fixedU_run_level <- 4
  vsenkf_run_level <- 2
}

if(is.numeric(run_level)){
  abfNbhd_run_level <- run_level
  lz_run_level <- run_level
  ij_run_level <- run_level
  bm_run_level <- run_level
  mscale_run_level <- run_level
  slice_run_level <- run_level
  fixedU_run_level <- run_level
  vsenkf_run_level <- run_level

if(run_level=="jasa-sept-2020"){  
#### values used for jasa submission, sept 2020  
  abfNbhd_run_level <- 3
  lz_run_level <- 5
  ij_run_level <- NA
  bm_run_level <- 4
  mscale_run_level <- 5
  slice_run_level <- 3
}
@

@@

\input{../header-ms.tex}
\input{../theorems.tex}

@@

%\author{Author1, Author2 and Author3 (Complete author list)}
%\correspondingauthor{Corresponding Author Name.\\E-mail: author.two@email.com}

\renewcommand{\contentsname}{Supplementary Content}
\renewcommand{\refname}{Supplementary References}
\renewcommand\thefigure{S-\arabic{figure}}
\renewcommand\thetable{S-\arabic{table}}
%\renewcommand\thepage{S-\arabic{page}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\theprop{S\arabic{prop}}
\renewcommand\thelemma{S\arabic{lemma}}
\renewcommand\thealgocf{S\arabic{algocf}}

@@

\begin{algorithm}[H]
  \caption{\bf Generalized unadapted bagged filter (G-UBF).
   }\label{alg:gubf}
  \KwIn{
    simulator for $f_{{X}_{\time}|{X}_{\time-1}}({x}_{\time}\given {x}_{\time-1})$ and $f_{{X}_0}({x}_0)$;
    evaluator for $f_{{Y}_{\unit,\time}|{X}_{\time}}({y}_{\unit,\time}\given {x}_{\time})$;
    data, $\data{\myvec{y}}_{1:\Time}$;
    number of replicates, $\Rep$;
    neighborhood structure, $B_{\unit,\time}$
  }
\For{$\rep\ \mathrm{in}\ \seq{1}{\Rep}$}{
simulate
  ${X}_{0:\Time,\rep}^{\simulation}$
  from the dynamic model, for $\time$ in $\seq{1}{\Time}$
  \;
prediction weights, 
  $w^P_{\unit,\time,\rep}= f_{Y_{B_{\unit,\time}}|{X}_{1:\time}}
    (\data{y}_{B_{\unit,\time}}\given {X}^{\simulation}_{1:\time,\rep})$
  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
  \;
measurement weights,
  $w^M_{\unit,\time,\rep}=f_{Y_{\unit,\time}|X_{\time},Y_{B_{\unit,\time}}}
    (\data{y}_{\unit,\time}\given
    X^{\simulation}_{\time,\rep},\data{y}_{B_{\unit,\time}})$
  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
}
  $\MC{\loglik}_{\unit,\time}= 
    \log \left(
      \sum_{\rep=1}^\Rep w^M_{\unit,\time,\rep}\,   w^P_{\unit,\time,\rep}
    \right)
    -
    \log \left(
        \sum_{\tilde \rep=1}^\Rep w^P_{\unit,\time,\tilde \rep}
    \right)$
  for $\unit$ in $\seq{1}{\Unit}$, $\time$ in $\seq{1}{\Time}$ 
\;
\KwOut{
log likelihood estimate, $\MC{\loglik}= \sum_{\time=1}^\Time\sum_{\unit=1}^\Unit \MC{\loglik}_{\unit\comma\time}$\\
}
\end{algorithm}

@@

Call the adapted simulation process $\{A_{\time},\time=1,2,\dots\}$, defined conditionally on $\{Y_{\time},\time=1,2,\dots\}$. 
We see from the above calculation that $A_{\time}$ can be constructed by the recursion

@@

$A_{\time}$ tracks $X_{\time}$ when the process $\{A_{\time}-X_{\time},\time=1,2,\dots\}$ is stable.

@@

Together, the results in \myeqref{thm1:loglik:linearization}, \myeqref{var:bound:without:assumption}, \myeqref{var:bound:with:assumption} and \myeqref{thm1:bias:bound} confirm the assertions of the theorem.

@@

The mixing of the adapted process in Assumption~\ref{B:unconditional:mix} replaces the mixing of the unconditional process in  Assumption~\ref{A:unconditional:mix}.
Though mixing of the adapted process may be hard to check, one may suspect that the adapted process typically mixes more rapidly than the unconditional process.
%As for the TIF analysis, we note that requiring the inequality in Assumption~\ref{B:unconditional:mix} to hold over ${B^{+2c}_{\unit,\time}}$ implies that it holds over all subsets of  ${B^{+2c}_{\unit,\time}}$. 
Assumption~\ref{B:unconditional:mix} is needed only to ensure that the variance bound in Theorem~\ref{thm:abf} is essentially $O(\Unit \Time)$ rather than $O(\Unit^2 \Time^2)$. 
Either of these rates avoids the exponentially increasing variance characterizing the curse of dimensionality.
Lower variance than $O(\Unit \Time)$ cannot be anticipated for any sequential Monte Carlo method since the log likelihood estimate can be written as a sum of $\Unit\Time$ terms each of which involves its own sequential Monte Carlo calculation.
The following Proposition~\ref{prop:cov:mix} gives an implication of Assumption~\ref{B:unconditional:mix}.

@@

bm_files_dir <- paste0("../bm_",bm_run_level,"/")
if(!dir.exists(bm_files_dir)) dir.create(bm_files_dir)

@@

measles_model_dir <- "../measlesModel/"
load(file=paste0(measles_model_dir,"measles_spatPomp.rda"))

@@

mscale_files_dir <- paste0("../mscale_",mscale_run_level,"/")

@@

\begin{table}
\renewcommand{\arraystretch}{1.2}
  \centering
  \begin{tabular}{|crcl|}
    \hline
    parameter & value & unit & description  \\  \hline
    $\meanBeta$ & \Sexpr{myround(measles_params["R0"]*(measles_params["gamma"]+measles_params["mu"]),1)} & year$^{-1}$& mean contact rate\\
    $\mu_{\bullet D}^{-1}$ & \Sexpr{myround(1/measles_params["mu"],1)} &year& mean duration in the population \\
    $\mu_{EI}^{-1}$ & \Sexpr{myround(365/measles_params["sigma"],1)} &day& latent period  \\
    $\mu_{IR}^{-1}$ & \Sexpr{myround(365/measles_params["gamma"],1)} &day& infectious period \\
    $\sigma_{SE}$ & \Sexpr{myround(measles_params["sigmaSE"],3)} &year$^{1/2}$& process noise\\
    $\amplitude$ & \Sexpr{myround(measles_params["amplitude"],3)} &--- & amplitude of seasonality  \\
    $\alpha$ & \Sexpr{measles_params["alpha"]} &---& mixing exponent \\
    $\tau$ & 4 & year & delay from birth to entry into susceptibles\\    
%%    $\cohort$ & \Sexpr{measles_params["amplitude"]} &---& cohort fraction entering $S$ on September 6 \\
    $\rho$ & \Sexpr{measles_params["rho"]} &---& reporting probability \\
    $\psi$ & \Sexpr{measles_params["psi"]} &---& reporting overdispersion \\
    $G$ & \Sexpr{measles_params["g"]} &---& gravitation constant \\
    $S_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["S1_0"]} &---& initial susceptible fraction \\
    $E_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["E1_0"]} &---& initial exposed fraction \\
    $I_{\unit}(0), \unit \in 1:\Unit$ & \Sexpr{measles_params["I1_0"]} &---& initial infectious fraction \\
\hline
  \end{tabular}
  \caption{\label{tab:measles_parameters} Parameters for the spatiotemporal measles transmission model}
\end{table}

@@

Table~\ref{tab:measles_parameters} gives the model parameter values.
The parameters were selected based on values found in time series analysis of single cities \citep{he10,bjornstad02}.
The coupling parameter, $G$, was chosen so that the number of cities with simulated epidemics outside the dominant 2-year cycle is visually similar to the data.
Initial values were chosen to initialize the system with the 2-year cycles observed in the UK during this period.
The model code was parameterized via the basic reproduction number, $\mathcal{R}_0=\bar\beta \big(\mu_{\bullet D}+\mu_{IR}\big)^{-1}$.
We used $\mathcal{R}_0=30$, corresponding to $\bar\beta = 1560.6 = 30 \times (52+0.02)$.
$\mathcal{R}_0=30$ is higher than most estimates for measles but consistent with estimates previously found via time series analysis of pre-vaccine UK measles in single cities \citep{he10,bjornstad02}.

Table~\ref{tab:mscale} gives the algorithmic settings used for the filters.
The times in Table~\ref{tab:mscale} give the total time required by each algorithm to calculate all its results for Figure~{\MainFigureMeaslesScaling} in the main text, using \Sexpr{mscale_cores} cores.
The expected forecast function $\mu(x,s,t)$ needed for {\ABFIR} and GIRF was computed using a numerical solution to the deterministic skeleton of the stochastic model, i.e, a system of ODEs with derivative matching the infinitesimal mean function of the stochastic dynamic model.
In the specifications of $h_\unit\big(x)$, $\VtoTheta_\unit(V,x,\theta)$ and $\thetaToV_\unit(\theta,x)$, the latent process value $x$ contains a variable $C$ giving the cumulative removed infections in the current observation interval.

In Table~\ref{tab:mscale}, we see that the effort allocated to {\UBF}, {\ABF} and PF scales linearly with $\Unit$, since the number of bootstrap replications and particles is fixed in this experiment. 
GIRF computational effort scales fairly linearly in $\Unit$, since its effort is dominated by the guide simulations (which are linear in $\Unit$) rather than by the intermediate timestep calculations (which are quadratic in $\Unit$ since we carry out $\Unit$ intermediate calculations each of size linear in $\Unit$).
The effort allocated to {\ABFIR} scales with $\Unit^2$, since {\ABFIR} is more parsimonious with guide simulations (all particles in one bootstrap replication share the same guide simulations) and so the intermediate timestep calculations dominate the effort.
To obtain stable variance in the log likelihood estimate, the number of particles and bootstrap replications would have to grow with $\Unit$.
However, given a constraint on total computational resources, the number of particles and bootstrap replications would have to shrink as $\Unit$ increases.
The limit studied in this experiment is a balance between the two: the assumption is that one is prepared to invest a growing amount of computational effort as the data grow, but this should not grow too fast.
{\ABFIR} was permitted the greatest computational effort, but the following two considerations balance this:
\begin{enumerate}
\item Parallelization. {\UBF}, {\ABF} and {\ABFIR} are trivially parallelizable. 
The value of parallelization depends, among other things, on how many replications are being computed simultaneously and on how many cores are available.
Nevertheless, it is helpful that the core minute effort requirement for {\ABF} and {\ABFIR} can be divided by the number of available cores to give the computational time. 
Parallelizations of GIRF and PF can be constructed \citep{park20} but these involve non-trivial interaction between processors leading to additional algorithmic complexity and computational overhead. 
\item Memory. 
The intermediate timestep calculations in {\ABFIR} and GIRF do not add to the memory requirement, and the memory demands of {\UBF}, {\ABF} and {\ABFIR} are distributed across the parallel computations. 
A basic PF implementation for a large model can become constrained by its memory requirement (linear in the number of particles) before it can match the processor effort employed by the other algorithms.
\end{enumerate}

@@

  abfNbhd_model_dir <- "../measlesModel/"
  # note: care is required if abfNbhd_model_dir is set to something other than the default measlesModel
  # however, it may be useful to do that while testing model variations.

@@

\section{\secTitleSpace A Lorenz-96 example}
\label{sec:lorenz}

@@

  set.seed(4512)
  lz_list <- foreach(u=lz_U) %do% {
    lz1 <- lorenz(
      U=u, N=lz_N,
      delta_t=lz_dt,delta_obs=lz_dt_obs,
      regular_params=lz_params)
    lz_ivps <- rnorm(u,mean=lz_ivp_mean,sd=lz_ivp_sd)
    names(lz_ivps) <- paste0("X",1:u,"_0")
    lz1_params <- coef(lz1)
    lz1_params[names(lz_ivps)] <- lz_ivps
    simulate(lz1,params=lz1_params)
  }
})

@@

<<lz_girf,cache=F,echo=F>>=
lz_girf <- stew(file=paste0(lz_files_dir,"lz_girf.rda"),seed=5981724,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      girf(lz_list[[job$U_id]], 
        Np=lz_girf_Np,
        Ninter = job$U,
	kind="moment",
        lookahead = lz_girf_lookahead,
        Nguide = lz_girf_nguide,
        tol = lz_tol) -> lz_girf_out 
    ) -> lz_girf_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"],lz_girf_out)
    list(logLik=logLik(lz_girf_out),time=lz_girf_time["elapsed"])
  } -> lz_girf_list
})
lz_jobs$girf_logLik <- vapply(lz_girf_list,function(x)x$logLik,numeric(1))
lz_jobs$girf_time <- vapply(lz_girf_list,function(x) x$time,numeric(1))
@

@@

<<lz_abfir,cache=F,echo=F,eval=T>>=
lz_abfir <- stew(file=paste0(lz_files_dir,"lz_abfir.rda"),seed=53398,{
  foreach(job=iter(lz_jobs,"row")) %do% {
    system.time(
      abfir(lz_list[[job$U_id]],
        Nrep = as.integer(lz_abfir_Nrep),
        Np=lz_abfir_Np_per_replicate,
        Ninter = as.integer(job$U/2),
        nbhd = lz_nbhd, tol=lz_tol) -> lz_abfir_out 
    ) -> lz_abfir_time
#    uncomment for debugging 
#    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"],lz_abfir_out)
    list(logLik=logLik(lz_abfir_out),time=lz_abfir_time["elapsed"])
  } -> lz_abfir_list
})
lz_jobs$abfir_logLik <- vapply(lz_abfir_list,function(x)x$logLik,numeric(1))
lz_jobs$abfir_time <- vapply(lz_abfir_list,function(x) x$time,numeric(1))
@

@@

lz_enkf <- stew(file=paste0(lz_files_dir,"lz_enkf.rda"),seed=837509,{
  foreach(job=iter(lz_jobs,"row")) %dopar% {
    system.time(
      enkf(lz_list[[job$U_id]], 
        Np=lz_enkf_Np) -> lz_enkf_out 
    ) -> lz_enkf_time
#    uncomment for debugging 
#    enkf(lz_list[[1]],Np=lz_enkf_Np)-> lztmp 
#    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"],lz_enkf_out)
    list(logLik=logLik(lz_enkf_out),time=lz_enkf_time["elapsed"])
  } -> lz_enkf_list
})
lz_jobs$enkf_logLik <- vapply(lz_enkf_list,function(x)x$logLik,numeric(1))
lz_jobs$enkf_time <- vapply(lz_enkf_list,function(x) x$time,numeric(1))

@@

<<lz_image_plot, echo=F, fig.height=3.5, fig.width=7, out.width="6.5in", fig.cap = paste('Lorenz \'96 simulation used for Figure \\ref{fig:lz_loglik_plot}')>>=
library(fields)
lz_obj <- lz_list[[which.max(lz_U)]]
par(mai=c(0.7,0.7,0.1,0.5))
image.plot(
  y=1:dim(states(lz_obj))[1],
  x=time(lz_obj),z=t(states(lz_obj)),
  ylab="unit",xlab="time",col=gray.colors(33)
)
@

@@

%%%%%%%%%%%%%%%%%%%%%% fls %%%%%%%%%%%%%%%%

\section{\secTitleSpace Bagged filters for functions of the latent states}

The theory and methodology in the main article focused on filtering for likelihood estimation.
Here, we describe extensions to other filtering problems.
Let $\{\filtFunc_{\filts}:\myvec{X}^{\Unit}\to\R, \mbox{ $\filts$ in $\seq{1}{\Filts}$}\}$ be a collection of functions where $\filtFunc_{\filts}$ depends only on a subset of units $\filtFuncSet_{\filts}\subset \seq{1}{\Unit}$.
We suppose that there exist neighborhoods
\begin{equation}
B^\prime_{\filts,\time}\subset A^\prime_{\time}=(\seq{1}{\Unit})\times(\seq{0}{\time})
\end{equation}
such that $\filtFunc_{\filts}(\myvec{X}_{\time})$ is approximately independent of $\{Y_{\unit,\time}:(\unit,\time) \in B^{\prime c}_{\filts,\time}\}$ given  $\{Y_{\unit,\time}: (\unit,\time) \in B^\prime_{\filts,\time}\}$, where $B^{\prime c}_{\filts,\time}$ is a complement in $A^\prime_{\time}$.
Unlike the sets $A_{\unit,\time}$ and $B_{\unit,\time}$ defined for likelihood estimation, the sets $A^\prime_{\time}$ and $B^\prime_{\filts,\time}$ can include any locations at time $\time$.
We consider Monte Carlo estimation of
\begin{equation}
\label{eq:filt1}
\overline{\filtFunc}_{k,\time}= \E\big[\filtFunc_{\filts}(\myvec{X}_\time) \big| \myvec{Y}_{1:\time}\big], \quad \filts \mbox{ in } \seq{1}{\Filts}.
\end{equation}
For example, if we set $\filtFunc_{\unit}(\myvec{x}_\time)=x_{\unit,\time}$ and $\Filts=\Unit$, the collection of quantities $\{\overline{\filtFunc}_{\unit,\time}\}$ estimated in \eqref{eq:filt1} corresponds to a vector of filter means.
Setting $\filtFunc_{\filts}(\myvec{x}_\time)=x_{\unit_{\filts},\time}\, x_{\altUnit_{\filts},\time}$ enables calculation of a collection of filter covariances between units $\unit_k$ and $\altUnit_k$ for $k$ in $\seq{1}{\Filts}$.

We consider bagged filtering approaches to estimation of $\{\overline{\filtFunc}_{\filts,\time}, \filts \mbox{ in } \seq{1}{\Filts}\}$.
Although each $\overline{\filtFunc}_{\filts,\time}$ is required to have only local dependence, some global quantities such as filter means and their variances across all units, can be expressed in terms of collections of such quantities.
For bagged filtering to operate successfully, the neighborhoods $\{B^\prime_{\filts,\time}\}$ should not be large.
Since  $B^\prime_{\filts,\time}$ will typically be larger than $\{\filtFuncSet_\filts\}$, this rules out estimation of filtered quantities that cannot be adequately represented by a collection of localized filtering calculations.
We now present variants of the pseudocode in the main text, targeted at estimation of $\{\overline{\filtFunc}_{k,\time},k \mbox{ in } 1{\mycolon}K\}$.
We do not prove theorems about these algorithms, but we conjecture from their similarity to the algorithms in the main text that comparable theoretical results should exist.
Table~\ref{tab:filter:inputs} lists the inputs, outputs and ranges of the implicit loops for the following algorithms.
